---
title: "ICPhS analysis of NZE diphthongs"
author: "Márton Sóskuthy"
date: "08/12/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Prep

Loading libraries...

```{r}
library(tidyverse)
library(phonTools)
library(gganimate)
library(mgcv)
library(ggforce)
library(parallel)
library(emuR)
library(directlabels)
```

Defining custom functions. First, a function for synthesising vowels.

```{r, include=F}
to_formant_track <- function (fname, f1bark, f2bark, dur, 
                              f3def=2800, f4def=3700, f5def=4600,
                              f6def=5500, f0_max=130, f0_min=110,
                              script_path=NULL,
                              praat_path=NULL) {
  f1 <- bark(f1bark, inv=T)
  f2 <- bark(f2bark, inv=T)
  f3 <- rep(f3def, length(f1))
  f4 <- rep(f4def, length(f1))
  f5 <- rep(f5def, length(f1))
  f6 <- rep(f6def, length(f1))
  locations <- seq(0,dur,length.out=length(f1))
  out <- data.frame(f1=f1, f2=f2, f3=f3, f4=f4, 
                    f5=f5, f6=f6, 
                    locations=locations)
  tempfname <- tempfile(fileext=".csv")
  # assume that script lives in ../praat/synthesise.praat
  if (is.null(script_path)) {
    script_path <- "../praat/synthesise.praat"
  }
  # assume that praat lives in "/Applications/Praat.app/Contents/MacOS/Praat"
  if (is.null(praat_path)) {
    praat_path <- "/Applications/Praat.app/Contents/MacOS/Praat"
  }
  write_csv(out, tempfname)
  system(paste0(praat_path, " --run ", script_path, ' "', tempfname, '" "', fname, '" ', f0_max, " ", f0_min))
  unlink(tempfname)
}
```

Data import.

```{r}
#price <- readRDS("../final_data/price_full.rds") # only average data are analysed so full data set not needed
price_avgs <- readRDS("../final_data/price_averaged.rds")
#mouth <- readRDS("../final_data/mouth_full.rds")
mouth_avgs <- readRDS("../final_data/mouth_averaged.rds")
#face <- readRDS("../final_data/face_full.rds")
face_avgs <- readRDS("../final_data/face_averaged.rds")
#goat <- readRDS("../final_data/goat_full.rds")
goat_avgs <- readRDS("../final_data/goat_averaged.rds")


# binding by-speaker avgs into single data set for price / mouth
price_avgs$vowel <- "price" 
mouth_avgs$vowel <- "mouth"
face_avgs$vowel <- "face"
goat_avgs$vowel <- "goat"
vowels_avgs <- rbind(price_avgs, mouth_avgs, face_avgs, goat_avgs)

# focus on pre-vd only, reformat so that data can be analysed using single model
vowels_avgs_vd <- filter(vowels_avgs, !as.logical(following_voiceless)) %>%
  gather(key="formant", value="hz", f1_avg, f2_avg, f3_avg) %>%
  filter(formant %in% c("f1_avg", "f2_avg"))
# variable for single model
vowels_avgs_vd$vowel_sex_formant <- interaction(vowels_avgs_vd$vowel, vowels_avgs_vd$sex, vowels_avgs_vd$formant)

# log duration variable
vowels_avgs_vd$dur_avg_log <- log(vowels_avgs_vd$dur_avg)

# focus on pre-vd only for main data
face <- readRDS("../final_data/face_full.rds")
goat <- readRDS("../final_data/goat_full.rds")
price <- readRDS("../final_data/price_full.rds")
mouth <- readRDS("../final_data/mouth_full.rds")

#price_vd <- filter(price, !following_voiceless)
#mouth_vd <- filter(mouth, !following_voiceless)
#face_vd <- filter(face, !following_voiceless)
#goat_vd <- filter(goat, !following_voiceless)
```

## GAM modelling of per-speaker avg contours

Fitting THE model. Note that vowel, sex and formant are *all* combined into a single factor (so each combination of vowel-sex-formant gets a completely independent smooth). Note that almost all model terms are significant. Note also that random effects by speaker are not needed, as there is only a single trajectory per speaker.

```{r}
# note that empirical rho = 0.877, so very close to the 0.9 used here
system.time({
  pm_f1f2_avg_mod <- 
    bam(hz ~ vowel_sex_formant +
          s(dur_avg_log, k=5, by=vowel_sex_formant) +
          s(measurement_no, k=11, by=vowel_sex_formant) + 
          s(yob, by=vowel_sex_formant, k=5) +
          ti(measurement_no, yob, k=c(11,5), by=vowel_sex_formant) +
          ti(measurement_no, dur_avg_log, k=c(11,5), by=vowel_sex_formant),
        data=vowels_avgs_vd, discrete=T, nthreads=2, 
        rho=0.9, AR.start=vowels_avgs_vd$AR_start)})
summary(pm_f1f2_avg_mod)
```

### Creating video summary

Video made simply via gganimate. Sound then added using custom synthesis function. Both males and females analysed here.

```{r}
# data frame for predictions
newdat <- expand.grid(measurement_no=seq(1,11,length.out=100),
                      yob=seq(min(vowels_avgs_vd$yob),
                              max(vowels_avgs_vd$yob),
                              1),
                      vowel_sex_formant=unique(vowels_avgs_vd$vowel_sex_formant),
                      dur_avg_log=median(vowels_avgs_vd$dur_avg_log),
                      hz=0)

#obtaining predictions from model
preds_hz <- predict(pm_f1f2_avg_mod, newdata=newdat,
                    se.fit=T)
newdat$hz <- preds_hz$fit

# splitting vowel_sex_formant into its components for ggplot
newdat$vowel <- unlist(lapply(as.character(newdat$vowel_sex_formant), function (x) strsplit(x, split="[.]")[[1]][1]))
newdat$vowel <- toupper(newdat$vowel)
newdat$sex <- unlist(lapply(as.character(newdat$vowel_sex_formant), function (x) strsplit(x, split="[.]")[[1]][2]))
newdat$sex <- ifelse(newdat$sex=='M', "male", "female")
newdat$formant <- unlist(lapply(as.character(newdat$vowel_sex_formant), function (x) strsplit(x, split="[.]")[[1]][3]))
newdat$measurement_no <- (newdat$measurement_no - 1)*10


for (v in unique(newdat$vowel)) {
#for (v in c("FACE","GOAT")) {
  for (s in unique(newdat$sex)) {
    # number of frames = number of years
    frame_no <- length(unique(newdat$yob))
    fps <- 10
    
    # dummy data set for highlighting in ggplot
    dd <- unique(select(newdat, vowel, sex))
    dd$colour <- ifelse(dd$vowel==v & dd$sex==s, "red", "x")
    
    # creating animated ggplot with one of the panels highlighted
    p <- ggplot(newdat, aes(x=measurement_no, y=hz, group=formant)) +
      geom_rect(data = dd, aes(fill = colour, x=NULL, y=NULL, group=NULL), xmin = -Inf,xmax = Inf,
        ymin = -Inf,ymax = Inf,alpha = 0.1) +
      scale_fill_manual(values=c("red","white")) +
      facet_grid(sex~vowel) +
      geom_line(lwd=2) +
      xlab("measurement point (% vowel duration)") +
      ylab("frequency (Bark)") +
      theme_bw() +
      theme(axis.title=element_text(size=18, face="bold"),
            axis.text=element_text(size=16),
            strip.text=element_text(size=16),
            plot.title=element_text(size=18, face="bold"),
            panel.grid=element_blank(),
            legend.position = "none") +
      transition_time(yob) +
      ease_aes('linear') +
      labs(title = 'Year of birth: {frame_time}')
    #animate(p, nframes=frame_no, fps=fps)
    
    # generating sound
    
    anim_dur <- frame_no/fps
    audio_samples <- 10
    audio_dur <- 0.25
    edge_window <- 0.2
    pause_dur <- (anim_dur - 2*edge_window - audio_samples*audio_dur) / (audio_samples-1)

    yob_min <- min(newdat$yob)
    yob_max <- max(newdat$yob)
    yob_range <- yob_max - yob_min
    synthesise_at_yobs <- round(yob_min + yob_range*(seq(edge_window, edge_window + (audio_samples-1)*(audio_dur + pause_dur), audio_dur + pause_dur)/anim_dur))

    dirname <- "../sounds/temp/"
    unlink(paste0(dirname, "*.wav"))
    vid.dirname <- "../videos/"
    outname <- "../sounds/temp/temp_continuum.wav"
    for (i in 1:length(synthesise_at_yobs)) {
      at_yob <- synthesise_at_yobs[i]
      f1bark <- filter(newdat, yob==at_yob, sex==s, vowel==v, formant=="f1_avg")$hz
      f2bark <- filter(newdat, yob==at_yob, sex==s, vowel==v, formant=="f2_avg")$hz
      if (s=="female") {
        to_formant_track(paste0(dirname, sprintf("temp%.3d.wav", i)), f1bark, f2bark, audio_dur, f3=3080, f4=4070, f5=5060, f6=6050, f0_max=210, f0_min=180)
      } else {
        to_formant_track(paste0(dirname, sprintf("temp%.3d.wav", i)), f1bark, f2bark, audio_dur)
      }
    }
    
    system(paste0("/Applications/Praat.app/Contents/MacOS/Praat --run ", "../praat/concatenate.praat ", '"', dirname, '" "', outname, '" ', edge_window, ' ', pause_dur))
    
    animate(p, nframes=frame_no, fps=fps, renderer=av_renderer(audio=outname, file=paste0(vid.dirname, v, "_", s, "_nwav_vid.mp4")), width=960, height=480)
  }
}
```

### A summary plot

Three x two plots, actually, to clearly show development over time. Plot series #1 shows formant trajectories for males (females?) at three different points: 1860, 1900, 1960. Plot series #2 shows formant trajectories in a vowel space (with corner vowels from h95).

```{r}
# data frame for predictions
newdat <- expand.grid(measurement_no=seq(1,11,length.out=500),
                      yob=c(1860, 1900, 1960),
                      vowel_sex_formant=unique(vowels_avgs_vd$vowel_sex_formant),
                      dur_avg_log=median(vowels_avgs_vd$dur_avg_log),
                      hz=0)

#obtaining predictions from model
preds_hz <- predict(pm_f1f2_avg_mod, newdata=newdat,
                    se.fit=T)
newdat$hz <- preds_hz$fit

# splitting vowel_sex_formant into its components for ggplot
newdat$vowel <- unlist(lapply(as.character(newdat$vowel_sex_formant), function (x) strsplit(x, split="[.]")[[1]][1]))
newdat$vowel <- toupper(newdat$vowel)
newdat$sex <- unlist(lapply(as.character(newdat$vowel_sex_formant), function (x) strsplit(x, split="[.]")[[1]][2]))
newdat$sex <- ifelse(newdat$sex=='M', "male", "female")
newdat$formant <- unlist(lapply(as.character(newdat$vowel_sex_formant), function (x) strsplit(x, split="[.]")[[1]][3]))
newdat$measurement_no <- (newdat$measurement_no - 1)*10

newdat <- newdat %>% 
  filter(sex=='male') %>%
  mutate(vowel=factor(vowel, levels=c("FACE","GOAT","PRICE","MOUTH")))

# creating ggplot
ggplot(newdat, aes(x=measurement_no, y=hz, group=formant, color=vowel)) +
  facet_grid(yob~vowel) +
  geom_line(lwd=1) +
  #scale_color_manual(values=c("#998ec3","#f1a340")) +
  scale_x_continuous(breaks=seq(0,100,25),
                     labels=c("", "25%", "50%", "75%", "")) +
  scale_color_viridis_d(guide=F) +
  xlab("measurement point (% vowel duration)") +
  ylab("frequency (Bark)") +
  theme_bw() +
  theme(axis.title=element_text(size=18, face="bold"),
        axis.text.y=element_text(size=14),
        axis.text.x=element_text(size=14, hjust=1, angle=45),
        strip.text=element_text(size=16),
        plot.title=element_text(size=18, face="bold"),
        panel.grid=element_blank(),
        strip.text.y=element_blank(),
        legend.position = "top",
        legend.title=element_text(size=16, face="bold"),
        legend.text=element_text(size=16))
ggsave("graphs/static_movie_nwav.pdf", width=7, height=7)

```

And now for series #2.

```{r}
library(phonTools)
data(h95)

newdat$vowel_sex <- unlist(lapply(str_split(newdat$vowel_sex_formant, "[.]"), function (x)
return(paste(x[1], x[2], sep="."))))
newdat_f1f2 <- newdat %>%
  dplyr::select(-vowel_sex_formant) %>%
  spread(formant, hz) %>%
  filter(sex=="male") %>%
  arrange(vowel, measurement_no) %>%
  mutate(vowel=factor(vowel, levels=c("FACE","GOAT","PRICE","MOUTH")))

corner_vs <- h95 %>%
  filter(vowel %in% c("i","u","A"), type=="m") %>%
  group_by(vowel) %>%
  summarise(f1bark=mean(bark(f1)),
            f2bark=mean(bark(f2)),
            f3bark=mean(bark(f3))) %>%
  ungroup() %>%
  mutate(vowel=recode(vowel, i="FLEECE", u="GOOSE", A="START"))

for (y in unique(newdat_f1f2$yob)) {
  ggplot(filter(newdat_f1f2, yob==y), aes(x=f2_avg, y=f1_avg, group=vowel, size=measurement_no, col=vowel)) +
    geom_path(lineend="round") +
    #geom_text(data=corner_vs, aes(x=f2bark, y=f1bark, label=vowel), col="black", size=5) +
    geom_point(data=corner_vs, aes(x=f2bark, y=f1bark), col="grey", size=1) +
    scale_color_viridis_d() +
    scale_x_reverse(limits=c(14.3,8)) +
    scale_y_reverse() +
    scale_alpha(guide=F) +
    scale_size(guide=F, range=c(0.2, 1.7)) +
    theme_bw() +
    theme(axis.title=element_blank(),
          axis.text=element_blank(),
          axis.ticks=element_blank(),
          axis.line=element_blank(),
          panel.border=element_blank(),
          panel.grid=element_blank(),
          strip.text.y=element_blank(),
          legend.position = "none")
  ggsave(paste0("graphs/vowel_chart_", y, ".pdf"), width=2.66, height=2)
}
```

## Making video for talk

So the idea is to create an animation with x=time y=F1/2 plots on the left and x=F2 y=F1 plots on the right with superimposed sound that reflects the grouping PRICE-MOUTH vs. FACE-GOAT. I'd also like to include some sort of markers to show 25%, 50% and 75% of vowel duration (e.g. notches?). Would also be nice to have labels for the vowels in the X=F2 plots. Best approach seems to be to create first animation (x=time) separately from second animation (x=F2), making sure that they are aligned by YOB; and then generate sound separately and add it on. Will be a wild ride!

```{r}
# LEFT PANEL

# data frame for predictions
newdat <- expand.grid(measurement_no=seq(1,11,length.out=50),
                      yob=seq(min(vowels_avgs_vd$yob),
                              max(vowels_avgs_vd$yob),
                              1),
                      vowel_sex_formant=unique(vowels_avgs_vd$vowel_sex_formant),
                      dur_avg_log=median(vowels_avgs_vd$dur_avg_log),
                      hz=0)

#obtaining predictions from model
preds_hz <- predict(pm_f1f2_avg_mod, newdata=newdat)
newdat$hz <- preds_hz

# splitting vowel_sex_formant into its components for ggplot
newdat$vowel <- unlist(lapply(as.character(newdat$vowel_sex_formant), function (x) strsplit(x, split="[.]")[[1]][1]))
newdat$vowel <- toupper(newdat$vowel)
newdat$sex <- unlist(lapply(as.character(newdat$vowel_sex_formant), function (x) strsplit(x, split="[.]")[[1]][2]))
newdat$sex <- ifelse(newdat$sex=='M', "male", "female")
newdat$formant <- unlist(lapply(as.character(newdat$vowel_sex_formant), function (x) strsplit(x, split="[.]")[[1]][3]))
newdat$measurement_no <- (newdat$measurement_no - 1)*10

newdat <- newdat %>% 
  mutate(vowel=as.ordered(factor(vowel, levels=c("PRICE","MOUTH", "FACE","GOAT"))),
         sex=factor(sex))

# creating ggplot
pleft <- ggplot(newdat, aes(x=measurement_no, y=hz, group=formant, color=vowel, frame=yob)) +
  facet_grid(rows=vars(sex), cols=vars(vowel), switch="y") +
  geom_line(lwd=1) +
  #scale_color_manual(values=c("#998ec3","#f1a340")) +
  scale_x_continuous(breaks=seq(0,100,25),
                     labels=c("", "25%", "50%", "75%", "")) +
  #scale_y_continuous(position="right") +
  scale_color_viridis_d(guide=F) +
  xlab("measurement point (% vowel duration)") +
  ylab("frequency (Bark)") +
  theme_bw() +
  theme(axis.title=element_text(size=18, face="bold"),
        axis.text.y=element_text(size=14),
        axis.text.x=element_text(size=14, hjust=1, angle=45),
        strip.text=element_text(size=16),
        plot.title=element_text(size=18, face="bold"),
        panel.grid=element_blank(),
        strip.placement="inside",
        legend.position = "none",
        plot.margin=margin(r=0.5, unit="npc")) +
  transition_time(yob) +
  ease_aes('linear') +
  labs(title = 'Year of birth: {frame_time}')
animate(pleft, nframes=length(unique(newdat$yob)), renderer=av_renderer(file="animations/all_changes_left.mp4"), width=960, height=480)
#animate(pleft, nframes=length(unique(newdat$yob)))
#ggsave("animations/left_panel.mov", width=7, height=7)

```

```{r}
frame_no = length(unique(newdat$yob))
fps = 10

# RIGHT PANEL

library(phonTools)
data(h95)

# data frame for predictions
newdat <- expand.grid(measurement_no=seq(1,11,length.out=50),
                      yob=seq(min(vowels_avgs_vd$yob),
                              max(vowels_avgs_vd$yob),
                              1),
                      vowel_sex_formant=unique(vowels_avgs_vd$vowel_sex_formant),
                      dur_avg_log=median(vowels_avgs_vd$dur_avg_log),
                      hz=0)

#obtaining predictions from model
preds_hz <- predict(pm_f1f2_avg_mod, newdata=newdat)
newdat$hz <- preds_hz

# splitting vowel_sex_formant into its components for ggplot
newdat$vowel <- unlist(lapply(as.character(newdat$vowel_sex_formant), function (x) strsplit(x, split="[.]")[[1]][1]))
newdat$vowel <- toupper(newdat$vowel)
newdat$sex <- unlist(lapply(as.character(newdat$vowel_sex_formant), function (x) strsplit(x, split="[.]")[[1]][2]))
newdat$sex <- ifelse(newdat$sex=='M', "male", "female")
newdat$formant <- unlist(lapply(as.character(newdat$vowel_sex_formant), function (x) strsplit(x, split="[.]")[[1]][3]))
newdat$measurement_no <- (newdat$measurement_no - 1)*10

#newdat$vowel_sex <- unlist(lapply(str_split(newdat$vowel_sex_formant, "[.]"), function (x) return(paste(x[1], x[2], sep="."))))
newdat_f1f2 <- newdat %>%
  dplyr::select(-vowel_sex_formant) %>%
  spread(formant, hz) %>%
  arrange(sex, vowel, measurement_no) %>%
  mutate(vowel=as.ordered(factor(vowel, levels=c("PRICE","MOUTH","FACE","GOAT"))))

newdat_f1f2_quantiles <- newdat_f1f2 %>%
  filter(measurement_no %in% quantile(unique(measurement_no), probs=c(0.25, 0.5, 0.75), type=1))
  


corner_vs <- h95 %>%
  filter(vowel %in% c("i","u","A"), type %in% c("m","w")) %>%
  group_by(vowel, type) %>%
  summarise(f1bark=mean(bark(f1)),
            f2bark=mean(bark(f2)),
            f3bark=mean(bark(f3))) %>%
  ungroup() %>%
  mutate(vowel=as.character(recode(vowel, i="i", u="u", A="a")),
         sex=recode(type, w="female", m="male"))

yobs <- rep(seq(min(newdat$yob), max(newdat$yob), 1), each=nrow(corner_vs))
corner_vs <- corner_vs[rep(1:nrow(corner_vs), times=length(unique(newdat$yob))),]
corner_vs$yob <- yobs

pright <- ggplot(newdat_f1f2, aes(x=f2_avg, y=f1_avg, group=vowel, size=measurement_no, col=vowel, frame=yob)) +
    facet_grid(rows=vars(sex)) +
    geom_path(lineend="round") +
    geom_point(data=newdat_f1f2_quantiles, size=2) +
    #geom_text(data=corner_vs, aes(x=f2bark, y=f1bark, label=vowel), col="black", size=5) +
    geom_text(data=corner_vs, aes(x=f2bark, y=f1bark, label=vowel), col="grey", size=8) +
    scale_color_viridis_d() +
    scale_x_reverse(limits=c(15.4,8)) +
    scale_y_reverse(limits=c(9.25,2.75)) +
    scale_alpha(guide=F) +
    scale_size(guide=F, range=c(0.2, 1.7)) +
    theme_bw() +
    theme(axis.title=element_blank(),
          axis.text=element_blank(),
          axis.ticks=element_blank(),
          axis.line=element_blank(),
          panel.border=element_blank(),
          panel.grid=element_blank(),
          strip.text.y=element_blank(),
          legend.position = "none") +
  transition_time(yob) +
  ease_aes('linear')# +
  #labs(title = 'Year of birth: {frame_time}')
animate(pright, nframes=length(unique(newdat$yob)), renderer=av_renderer(file="animations/all_changes_right.mp4"), width=240, height=480)
#animate(pright, nframes=)
#ggsave(paste0("graphs/vowel_chart_", y, ".pdf"), width=2.66, height=2)
```

Creating the corresponding sound.

```{r}
# generating sound - all
    
anim_dur <- frame_no/fps
audio_samples <- 5
audio_dur <- 0.27
edge_window <- 0.2
short_break_dur <- 0.25
total_dur <- audio_dur*4 + short_break_dur*3

pause_dur <- (anim_dur - 2*edge_window - audio_samples*total_dur) / (audio_samples-1)


yob_min <- min(newdat$yob)
yob_max <- max(newdat$yob)
yob_range <- yob_max - yob_min
synthesise_at_yobs <- round(yob_min + yob_range*(seq(edge_window, edge_window + (audio_samples-1)*(total_dur + pause_dur), total_dur + pause_dur)/anim_dur))

dirname <- "/users/soskuthy/documents/research/current/2018/nz_vowels/nwav/animations/temp_sounds/"

#vid.dirname <- "../videos/"
outname_dir <- "/users/soskuthy/documents/research/current/2018/nz_vowels/nwav/animations/temp_sounds/"
final_outname <- "/users/soskuthy/documents/research/current/2018/nz_vowels/nwav/animations/all_Vs_continuum.wav"
s <- "female"
vowels <- c("PRICE","MOUTH","FACE","GOAT")

for (i in 1:length(synthesise_at_yobs)) {
  v_num <- 0
  unlink(paste0(dirname, i, "/", "*.wav"))  
  for (v in vowels) {
    v_num <- v_num + 1 
    at_yob <- synthesise_at_yobs[i]
    f1bark <- filter(newdat, yob==at_yob, sex==s, vowel==v, formant=="f1_avg")$hz
    f2bark <- filter(newdat, yob==at_yob, sex==s, vowel==v, formant=="f2_avg")$hz
    if (s=="female") {
      to_formant_track(paste0(dirname, i, "/", sprintf("temp%.3d.wav", v_num)), f1bark, f2bark, audio_dur, f3=2950, f4=3850, f5=4850, f6=5800, f0_max=250, f0_min=200)
    } else {
      to_formant_track(paste0(dirname, i, "/", sprintf("temp%.3d.wav", v_num)), f1bark, f2bark, audio_dur)
    }
  }
}

for (i in 1:audio_samples) {
  outname <- paste0(outname_dir, "vowel_block_", i, ".wav")
  system(paste0("/Applications/Praat.app/Contents/MacOS/Praat --run ", "../praat/concatenate.praat ", '"', dirname, i, "/", '" "', outname, '" ', 0.001, ' ', short_break_dur))
}

system(paste0("/Applications/Praat.app/Contents/MacOS/Praat --run ", "../praat/concatenate.praat ", '"', dirname, '" "', final_outname, '" ', edge_window, ' ', pause_dur))
```

## Single-point video

```{r}
# RIGHT PANEL

library(phonTools)
data(h95)

# data frame for predictions
newdat <- expand.grid(measurement_no=4,
                      yob=seq(min(vowels_avgs_vd$yob),
                              max(vowels_avgs_vd$yob),
                              1),
                      vowel_sex_formant=unique(vowels_avgs_vd$vowel_sex_formant),
                      dur_avg_log=median(vowels_avgs_vd$dur_avg_log),
                      hz=0)

frame_no = length(unique(newdat$yob))
fps = 10

#obtaining predictions from model
preds_hz <- predict(pm_f1f2_avg_mod, newdata=newdat)
newdat$hz <- preds_hz

# splitting vowel_sex_formant into its components for ggplot
newdat$vowel <- unlist(lapply(as.character(newdat$vowel_sex_formant), function (x) strsplit(x, split="[.]")[[1]][1]))
newdat$vowel <- toupper(newdat$vowel)
newdat$sex <- unlist(lapply(as.character(newdat$vowel_sex_formant), function (x) strsplit(x, split="[.]")[[1]][2]))
newdat$sex <- ifelse(newdat$sex=='M', "male", "female")
newdat$formant <- unlist(lapply(as.character(newdat$vowel_sex_formant), function (x) strsplit(x, split="[.]")[[1]][3]))
newdat$measurement_no <- (newdat$measurement_no - 1)*10

#newdat$vowel_sex <- unlist(lapply(str_split(newdat$vowel_sex_formant, "[.]"), function (x) return(paste(x[1], x[2], sep="."))))
newdat_f1f2 <- newdat %>%
  dplyr::select(-vowel_sex_formant) %>%
  spread(formant, hz) %>%
  arrange(sex, vowel, measurement_no) %>%
  mutate(vowel=as.ordered(factor(vowel, levels=c("PRICE","MOUTH","FACE","GOAT"))))

corner_vs <- h95 %>%
  filter(vowel %in% c("i","u","A"), type %in% c("m","w")) %>%
  group_by(vowel, type) %>%
  summarise(f1bark=mean(bark(f1)),
            f2bark=mean(bark(f2)),
            f3bark=mean(bark(f3))) %>%
  ungroup() %>%
  mutate(vowel=as.character(recode(vowel, i="i", u="u", A="a")),
         sex=recode(type, w="female", m="male"))

yobs <- rep(seq(min(newdat$yob), max(newdat$yob), 1), each=nrow(corner_vs))
corner_vs <- corner_vs[rep(1:nrow(corner_vs), times=length(unique(newdat$yob))),]
corner_vs$yob <- yobs

single_point <- ggplot(newdat_f1f2, aes(x=f2_avg, y=f1_avg, col=vowel, frame=yob)) +
    facet_grid(cols=vars(sex)) +
    geom_label(aes(label=vowel), size=7) +
    #geom_text(data=corner_vs, aes(x=f2bark, y=f1bark, label=vowel), col="black", size=5) +
    geom_text(data=corner_vs, aes(x=f2bark, y=f1bark, label=vowel), col="grey", size=8) +
    scale_color_viridis_d() +
    scale_x_reverse(limits=c(15.4,8)) +
    scale_y_reverse(limits=c(9.25,2.75)) +
    theme_bw() +
    theme(axis.title=element_blank(),
          axis.text=element_blank(),
          axis.ticks=element_blank(),
          axis.line=element_blank(),
          panel.border=element_blank(),
          panel.grid=element_blank(),
          strip.text.x=element_text(size=18, face="bold"),
          strip.background=element_blank(),
          legend.position = "none",
          plot.title=element_text(size=18),
          panel.background=element_rect(fill="#EAF0EF"),
          plot.background=element_rect(fill="#EAF0EF")) +
  transition_time(yob) +
  ease_aes('linear') +
  labs(title = 'Year of birth: {frame_time}')
animate(single_point, nframes=length(unique(newdat$yob)), renderer=av_renderer(file="animations/single_point_changes.mp4"), width=720, height=480)
#animate(pright, nframes=)
#ggsave(paste0("graphs/vowel_chart_", y, ".pdf"), width=2.66, height=2)
```