---
title: "ICPhS analysis of NZE diphthongs"
author: "Márton Sóskuthy"
date: "08/12/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Prep

Loading libraries...

```{r}
library(tidyverse)
library(phonTools)
library(gganimate)
library(mgcv)
library(ggforce)
library(parallel)
library(emuR)
library(directlabels)
```

Defining custom functions. First, a function for synthesising vowels.

```{r, include=F}
to_formant_track <- function (fname, f1bark, f2bark, dur, 
                              f3def=2800, f4def=3700, f5def=4600,
                              f6def=5500, f0_max=130, f0_min=110,
                              script_path=NULL,
                              praat_path=NULL) {
  f1 <- bark(f1bark, inv=T)
  f2 <- bark(f2bark, inv=T)
  f3 <- rep(f3def, length(f1))
  f4 <- rep(f4def, length(f1))
  f5 <- rep(f5def, length(f1))
  f6 <- rep(f6def, length(f1))
  locations <- seq(0,dur,length.out=length(f1))
  out <- data.frame(f1=f1, f2=f2, f3=f3, f4=f4, 
                    f5=f5, f6=f6, 
                    locations=locations)
  tempfname <- tempfile(fileext=".csv")
  # assume that script lives in ../praat/synthesise.praat
  if (is.null(script_path)) {
    script_path <- "../praat/synthesise.praat"
  }
  # assume that praat lives in "/Applications/Praat.app/Contents/MacOS/Praat"
  if (is.null(praat_path)) {
    praat_path <- "/Applications/Praat.app/Contents/MacOS/Praat"
  }
  write_csv(out, tempfname)
  system(paste0(praat_path, " --run ", script_path, ' "', tempfname, '" "', fname, '" ', f0_max, " ", f0_min))
  unlink(tempfname)
}
```

Second, diabolically complicated function for identifying points of maximum convexity / concavity in trajectories.

```{r, include=F}
find_inflection <- function (xs, ys, k, groups, convex=T, n=1000, plot=F, .pb=NULL) {
  if ((!is.null(.pb)) && inherits(.pb, "Progress") && (.pb$i < .pb$n)) 
    .pb$tick()$print()
  d <- data.frame(xs, ys, groups)
  d <- d[order(groups, xs),]
  d$start <- d$xs == min(d$xs)
  d$groups <- as.factor(d$groups)
  if (length(unique(groups)) < 3) {
    warning(paste0('too few trajectories (< 3), returning NA'))
    return(NA)
  }
  if (k > length(unique(xs))) {
    if (length(unique(xs)) < 9) {
      warning(paste0('trajectories have too many missing points (more > 2), returning NA'))
      return(NA)
    }
    k <- length(unique(xs))
    warning(paste0('k reduced to ', k))
  }
  
  # fit gamm with AR1 autoregressive error model
  modAR <- tryCatch({
    gamm(ys ~ s(xs, k=k, bs="cr"),
              data=d, correlation=corAR1(form=~ xs | groups))
    },
    error=function (x) NULL
  )
  if (is.null(modAR)) {
    warning("model could not be fit to trajectories")
    return(NA)
  }
  
  new_xs <- seq(min(d$xs), max(d$xs), length.out=n)
  newdat <- data.frame(xs=new_xs,
                       ys=0)
  
  # get first derivative of smooth
  eps <- 1e-04
  X0 <- predict(modAR$gam, newdat, type = 'lpmatrix')
  newdat_eps_1 <- newdat
  newdat_eps_1$xs <- newdat_eps_1$xs + eps
  X1 <- predict(modAR$gam, newdat_eps_1, type = 'lpmatrix')
  
  # finite difference approximation of first derivative
  # the design matrix
  Xp <- (X1 - X0) / eps
  first_deriv <- Xp %*% coef(modAR$gam)
  
  # get second derivative
  newdat_eps_2 <- newdat
  newdat_eps_2$xs <- newdat_eps_2$xs - eps
  X_1 <- predict(modAR$gam, newdat_eps_2, type = 'lpmatrix')
  # design matrix for second derivative
  Xpp <- (X1 + X_1 - 2*X0)  / eps^2
  # second derivative
  second_deriv <- Xpp %*% coef(modAR$gam)
  smooth_second_deriv <- predict(gamm(second_deriv~s(new_xs))$gam)
  
  # 1) does first derivative pass through 0? where?
  
  fd_shift_1 <- first_deriv[-length(first_deriv)]
  fd_shift_2 <- first_deriv[-1]
  xings <- fd_shift_1 * fd_shift_2 <= 0
  can_return <- F
  if (sum(xings) > 0) {
    indices <- (1:length(xings))[xings]
    d2s <- smooth_second_deriv[indices]
    if (convex & any(d2s > 0)) {
      out <- new_xs[indices[which.max(d2s)]]
      can_return <- T
    } else if (!convex & any(d2s < 0)) {
      out <- new_xs[indices[which.min(d2s)]]
      can_return <- T
    }
  } 
  if (!can_return) {
  
  # 2) ok, didn't find anything, now return 
  #    point of maximum convexity / concavity 
  # (max / min of second derivative, respectively)
  
    if (convex) {
      out <- new_xs[which.max(smooth_second_deriv)]
    } else {
      out <- new_xs[which.min(smooth_second_deriv)]
    }
  }
  
  if (plot) {
    par(mfrow=c(3,1))
    par(mar=c(2.1, 4.1, 0.1, 0.1))
    plot(new_xs, predict(modAR$gam, newdata=newdat))
    lines(1:11, aggregate(ys ~ xs, FUN=mean)$ys, col="red")
    abline(v=out, col="red")
    plot(new_xs, first_deriv)
    plot(new_xs, smooth_second_deriv)
    lines(new_xs, second_deriv, col="red")
  }
  
  return(out)
}

# example for usage:
 #spe <- filter(price, speaker=="fop96-2b", following_voiceless==T)
 #xs <- spe$measurement_no; ys <- spe$f2; k=10; groups=spe$id; n=1000
 #find_inflection(xs, ys, k, groups, convex=T, plot=T)
```

Data import.

```{r}
price <- readRDS("../final_data/price_full.rds")
price_avgs <- readRDS("../final_data/price_averaged.rds")
mouth <- readRDS("../final_data/mouth_full.rds")
mouth_avgs <- readRDS("../final_data/mouth_averaged.rds")

# binding by-speaker avgs into single data set for price / mouth
price_avgs$vowel <- "price" 
mouth_avgs$vowel <- "mouth"
price_mouth_avgs <- rbind(price_avgs, mouth_avgs)

# focus on pre-vd only, reformat so that data can be analysed using single model
price_mouth_avgs_vd <- filter(price_mouth_avgs, !as.logical(following_voiceless)) %>%
  gather(key="formant", value="hz", f1_avg, f2_avg, f3_avg) %>%
  filter(formant %in% c("f1_avg", "f2_avg"))
# variable for single model
price_mouth_avgs_vd$vowel_sex_formant <- interaction(price_mouth_avgs_vd$vowel, price_mouth_avgs_vd$sex, price_mouth_avgs_vd$formant)

# log duration variable
price_mouth_avgs_vd$dur_avg_log <- log(price_mouth_avgs_vd$dur_avg)

# focus on pre-vd only for main data
price_vd <- filter(price, !following_voiceless)
mouth_vd <- filter(mouth, !following_voiceless)
```

## GAM modelling of per-speaker avg contours

Fitting THE model.

```{r}
system.time({
  pm_f1f2_avg_mod <- 
    bam(hz ~ vowel_sex_formant +
          s(dur_avg_log, k=5, by=vowel_sex_formant) +
          s(measurement_no, k=11, by=vowel_sex_formant) + 
          s(yob, by=vowel_sex_formant, k=5) +
          ti(measurement_no, yob, k=c(11,5), by=vowel_sex_formant) +
          ti(measurement_no, dur_avg_log, k=c(11,5), by=vowel_sex_formant),
        data=price_mouth_avgs_vd, discrete=T, nthreads=2, 
        rho=0.9, AR.start=price_mouth_avgs_vd$AR_start)})
summary(pm_f1f2_avg_mod)
```

### Creating video summary

```{r}
# data frame for predictions
newdat <- expand.grid(measurement_no=seq(1,11,length.out=100),
                      yob=seq(min(price_mouth_avgs_vd$yob),
                              max(price_mouth_avgs_vd$yob),
                              1),
                      vowel_sex_formant=unique(price_mouth_avgs_vd$vowel_sex_formant),
                      dur_avg_log=median(price_mouth_avgs_vd$dur_avg_log),
                      hz=0)

#obtaining predictions from model
preds_hz <- predict(pm_f1f2_avg_mod, newdata=newdat,
                    se.fit=T)
newdat$hz <- preds_hz$fit

# splitting vowel_sex_formant into its components for ggplot
newdat$vowel <- unlist(lapply(as.character(newdat$vowel_sex_formant), function (x) strsplit(x, split="[.]")[[1]][1]))
newdat$vowel <- ifelse(newdat$vowel=='price', "PRICE", "MOUTH")
newdat$sex <- unlist(lapply(as.character(newdat$vowel_sex_formant), function (x) strsplit(x, split="[.]")[[1]][2]))
newdat$sex <- ifelse(newdat$sex=='M', "male", "female")
newdat$formant <- unlist(lapply(as.character(newdat$vowel_sex_formant), function (x) strsplit(x, split="[.]")[[1]][3]))
newdat$measurement_no <- (newdat$measurement_no - 1)*10


for (v in unique(newdat$vowel)) {
  for (s in unique(newdat$sex)) {
    # number of frames = number of years
    frame_no <- length(unique(newdat$yob))
    fps <- 10
    
    # dummy data set for highlighting in ggplot
    dd <- unique(select(newdat, vowel, sex))
    dd$colour <- ifelse(dd$vowel==v & dd$sex==s, "red", "x")
    
    # creating animated ggplot with one of the panels highlighted
    p <- ggplot(newdat, aes(x=measurement_no, y=hz, group=formant)) +
      geom_rect(data = dd, aes(fill = colour, x=NULL, y=NULL, group=NULL), xmin = -Inf,xmax = Inf,
        ymin = -Inf,ymax = Inf,alpha = 0.1) +
      scale_fill_manual(values=c("red","white")) +
      facet_grid(sex~vowel) +
      geom_line(lwd=2) +
      xlab("measurement point (% vowel duration)") +
      ylab("frequency (Bark)") +
      theme_bw() +
      theme(axis.title=element_text(size=18, face="bold"),
            axis.text=element_text(size=16),
            strip.text=element_text(size=16),
            plot.title=element_text(size=18, face="bold"),
            panel.grid=element_blank(),
            legend.position = "none") +
      transition_time(yob) +
      ease_aes('linear') +
      labs(title = 'Year of birth: {frame_time}')
    #animate(p, nframes=frame_no, fps=fps)
    
    # generating sound
    
    anim_dur <- frame_no/fps
    audio_samples <- 10
    audio_dur <- 0.25
    edge_window <- 0.2
    pause_dur <- (anim_dur - 2*edge_window - audio_samples*audio_dur) / (audio_samples-1)

    yob_min <- min(newdat$yob)
    yob_max <- max(newdat$yob)
    yob_range <- yob_max - yob_min
    synthesise_at_yobs <- round(yob_min + yob_range*(seq(edge_window, edge_window + (audio_samples-1)*(audio_dur + pause_dur), audio_dur + pause_dur)/anim_dur))

    dirname <- "../sounds/temp/"
    unlink(paste0(dirname, "*.wav"))
    vid.dirname <- "../videos/"
    outname <- "../sounds/temp/temp_continuum.wav"
    for (i in 1:length(synthesise_at_yobs)) {
      at_yob <- synthesise_at_yobs[i]
      f1bark <- filter(newdat, yob==at_yob, sex==s, vowel==v, formant=="f1_avg")$hz
      f2bark <- filter(newdat, yob==at_yob, sex==s, vowel==v, formant=="f2_avg")$hz
      if (s=="female") {
        to_formant_track(paste0(dirname, sprintf("temp%.3d.wav", i)), f1bark, f2bark, audio_dur, f3=3080, f4=4070, f5=5060, f6=6050, f0_max=210, f0_min=180)
      } else {
        to_formant_track(paste0(dirname, sprintf("temp%.3d.wav", i)), f1bark, f2bark, audio_dur)
      }
    }
    
    system(paste0("/Applications/Praat.app/Contents/MacOS/Praat --run ", "../praat/concatenate.praat ", '"', dirname, '" "', outname, '" ', edge_window, ' ', pause_dur))
    
    animate(p, nframes=frame_no, fps=fps, renderer=av_renderer(audio=outname, file=paste0(vid.dirname, v, "_", s, "_vid.mp4")))
  }
}
```

### A summary plot 

```{r}
# data frame for predictions
newdat <- expand.grid(measurement_no=seq(1,11,length.out=100),
                      yob=c(1860, 1900, 1960, 1985),
                      vowel_sex_formant=unique(price_mouth_avgs_vd$vowel_sex_formant),
                      dur_avg_log=median(price_mouth_avgs_vd$dur_avg_log),
                      hz=0)

#obtaining predictions from model
preds_hz <- predict(pm_f1f2_avg_mod, newdata=newdat,
                    se.fit=T)
newdat$hz <- preds_hz$fit

# splitting vowel_sex_formant into its components for ggplot
newdat$vowel <- unlist(lapply(as.character(newdat$vowel_sex_formant), function (x) strsplit(x, split="[.]")[[1]][1]))
newdat$vowel <- ifelse(newdat$vowel=='price', "PRICE", "MOUTH")
newdat$sex <- unlist(lapply(as.character(newdat$vowel_sex_formant), function (x) strsplit(x, split="[.]")[[1]][2]))
newdat$sex <- ifelse(newdat$sex=='M', "male", "female")
newdat$formant <- unlist(lapply(as.character(newdat$vowel_sex_formant), function (x) strsplit(x, split="[.]")[[1]][3]))
newdat$measurement_no <- (newdat$measurement_no - 1)*10


# creating ggplot
ggplot(newdat, aes(x=measurement_no, y=hz, group=interaction(formant, sex), col=sex)) +
  facet_grid(vowel~yob) +
  geom_line(lwd=1) +
  scale_color_manual(values=c("#998ec3","#f1a340")) +
  scale_x_continuous(breaks=seq(0,100,25),
                     labels=c("", "25%", "50%", "75%", "")) +
  xlab("measurement point (% vowel duration)") +
  ylab("frequency (Bark)") +
  theme_bw() +
  theme(axis.title=element_text(size=18, face="bold"),
        axis.text.y=element_text(size=16),
        axis.text.x=element_text(size=16, hjust=1, angle=25),
        strip.text=element_text(size=16),
        plot.title=element_text(size=18, face="bold"),
        panel.grid=element_blank(),
        legend.position = "top",
        legend.title=element_text(size=16, face="bold"),
        legend.text=element_text(size=16))
ggsave("graphs/static_movie.pdf", width=10, height=5)

```

## Inflection point analysis

Automatically detecting inflection points for each smooth fitted to each combination of speaker / vowel / formant.

```{r}
# set pb to no. of groups * 2
#pb = progress_estimated(length(unique(price_vd$speaker)) * 2)
price_vd_infl <- price_vd %>%
  group_by(speaker) %>%
  summarise(dur_avg=mean(dur),
            f1_infl=find_inflection(measurement_no, 
                                      bark(f1),
                                      k=10,
                                      groups=id,
                                      plot=F, convex=F),
            f2_infl=find_inflection(measurement_no, 
                                      bark(f2),
                                      k=10,
                                      groups=id,
                                      plot=F, convex=T),
            yob=yob[1], sex=sex[1]) %>%
  ungroup()
pb = progress_estimated(length(unique(mouth_vd$speaker)) * 2)
mouth_vd_infl <- mouth_vd %>%
  group_by(speaker) %>%
  summarise(dur_avg=mean(dur),
            f1_infl=find_inflection(measurement_no, 
                                       bark(f1),
                                       k=10,
                                       groups=id,
                                       plot=F, convex=F,
                                       .pb=pb),
            f2_infl=find_inflection(measurement_no, 
                                      bark(f2),
                                      k=10,
                                      groups=id,
                                      plot=F, convex=F,
                                      .pb=pb),
            yob=yob[1], sex=sex[1]) %>%
  ungroup()

# setting up for single gam analysis
mouth_vd_infl <- mouth_vd_infl %>%
  mutate(vowel="mouth") #%>%
  #rename(f1_infl="f1_concave",
  #       f2_infl="f2_concave")
price_vd_infl <- price_vd_infl %>%
  mutate(vowel="price") #%>%
  #rename(f1_infl="f1_concave",
  #       f2_infl="f2_convex")
pm_infl <- rbind(mouth_vd_infl, price_vd_infl) %>%
  gather(key="formant", value="inflection_point", c("f1_infl", "f2_infl"))
pm_infl$vowel_sex_formant <- interaction(pm_infl$vowel, pm_infl$sex, pm_infl$formant)
pm_infl$dur_avg_log <- log(pm_infl$dur_avg)
pm_infl$speaker <- as.factor(pm_infl$speaker)
```

And now for some GAM modelling!

```{r}
pm_infl_mod <- bam(inflection_point ~ vowel_sex_formant + 
                     s(dur_avg_log, k=5) +
                     s(yob, by=vowel_sex_formant, k=5) +
                     s(speaker, bs="re"),
                   data=pm_infl)
# who's SIGNIFICANT??
summary(pm_infl_mod)
```

Summary: all F1 timing changes are sig (and survive & even flourish after Bonferroni correction), but only male PRICE F2 is significant out of F2 timing changes. But don't read too much into significance due to exploratory nature of these tests. However, the patterns look extremely robust and correspond well with our visual impressions from the video / prediction plot from the previous section.

Generating prediction plot

```{r}

newdat <- expand.grid(yob=seq(min(pm_infl$yob),max(pm_infl$yob),1),
                      vowel_sex_formant=unique(pm_infl$vowel_sex_formant),
                      dur_avg_log=median(pm_infl$dur_avg_log),
                      speaker=pm_infl$speaker[1],
                      inflection_point=0)


#obtaining predictions from model
infl_preds <- predict(pm_infl_mod, newdata=newdat, exclude="s(speaker)",
                    se.fit=T)
newdat$inflection_point <- infl_preds$fit
newdat$lower <- infl_preds$fit - 1.96*infl_preds$se.fit
newdat$upper <- infl_preds$fit + 1.96*infl_preds$se.fit

# splitting vowel_sex_formant into its components for ggplot
newdat$vowel <- unlist(lapply(as.character(newdat$vowel_sex_formant), function (x) strsplit(x, split="[.]")[[1]][1]))
newdat$vowel <- ifelse(newdat$vowel=='price', "PRICE", "MOUTH")
newdat$sex <- unlist(lapply(as.character(newdat$vowel_sex_formant), function (x) strsplit(x, split="[.]")[[1]][2]))
newdat$sex <- ifelse(newdat$sex=='M', "male", "female")
newdat$formant <- unlist(lapply(as.character(newdat$vowel_sex_formant), function (x) strsplit(x, split="[.]")[[1]][3]))
newdat$formant <- ifelse(newdat$formant=='f1_infl', "F1", "F2")
newdat$sig <- ifelse(newdat$formant=="F1" | (newdat$formant=="F2" & newdat$vowel=="PRICE" & newdat$sex=="male"), 1, 0.6)
newdat$inflection_point <- newdat$inflection_point*10
newdat$lower <- newdat$lower*10
newdat$upper <- newdat$upper*10

# labels for the plot

labels <- data.frame(
  vowel=c("PRICE","MOUTH","PRICE","MOUTH","PRICE","MOUTH","PRICE","MOUTH"),
  sex=c("female","female","male","male","female","female","male","male"),
  formant=c("F1","F1","F1","F1","F2","F2","F2","F2"),
  sig=c(1,1,1,1,0.6,0.6,1,0.6),
  yob=c(1930,1910,1930,1910,1930,1930,1930,1930),
  inflection_point=c(5.5,8,5.5,8,5.2,2.7,5,2.7)*10
)

# creating ggplot
ggplot(newdat, aes(x=yob, y=inflection_point, group=vowel, col=vowel, alpha=sig)) +
  facet_grid(sex~formant) +
  geom_ribbon(aes(ymin=lower, ymax=upper), col=NA, fill="grey", alpha=0.3) +
  geom_line(lwd=1) +
  scale_color_manual(values=c("deepskyblue4","firebrick3")) +
  geom_text(data=labels, aes(label=vowel)) +
  #geom_dl(aes(label=vowel), method="lines2") +
  scale_y_continuous(breaks=seq(20,80,20),
                     labels=c("20%", "40%", "60%", "80%")) +
  xlab("year of birth") +
  ylab("inflection point (% V duration)") +
  theme_bw() +
  theme(axis.title=element_text(size=18, face="bold"),
        axis.text.y=element_text(size=16),
        axis.text.x=element_text(size=16),# hjust=1, angle=25),
        strip.text=element_text(size=16),
        plot.title=element_text(size=18, face="bold"),
        panel.grid=element_blank(),
        legend.position = "none")
ggsave("graphs/inflection_points.pdf", width=5, height=5)
```